---
format: gfm
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig-align: center
    fig-path: "man/figures/README-"
    fig-format: svg
---

<!-- README.md is generated from README.qmd. Please edit that file -->

# extera

<!-- badges: start -->
[![R-CMD-check](https://github.com/kbvernon/extera/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kbvernon/extera/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/extera)](https://CRAN.R-project.org/package=extera)
[![extendr](https://img.shields.io/badge/extendr-^0.8.0-276DC2)](https://extendr.github.io/extendr/extendr_api/)
<!-- badges: end -->

The name `extera` is a portmandeau of `extendr` and `tera`, making it suggestive
of the package's intended purpose, which is to provide an 
[`extendr`](https://github.com/extendr/extendr)-powered R wrapper around the 
blazing fast [`tera`](https://github.com/Keats/tera) templating-engine in Rust.

## Installation

You can install the development version of extera like so:

``` r
# install.packages("pak")
pak::pak("kbvernon/extera")
```

## Example

Everything in `extera` revolves around the `ExTera` object. You can initialize
an `ExTera` with an empty template library simply by calling `ExTera$new()`.

```{r}
#| label: basic-example
library(extera)

tera <- ExTera$new()

tera$add_string_templates(
  "hello-world" = '<p>Hello {{ x }}. This is {{ y }}.</p>',
  "star-wars" = '<h2>{{ title }}</h2>
<ol>
{%- for person in people %}
  {%- if person.films is containing("A New Hope") %}
  {%- if person.species and person.species is containing("Human") %}
  <li>{{ person.name }} ({{ person.homeworld }})</li>
  {%- endif %}
  {%- endif %}
{%- endfor %}
</ol>
'
)

tera

starwars <- dplyr::starwars[c("name", "films", "homeworld", "species")]

tera$render_to_string(
  "star-wars",
  title = "Humans of A New Hope",
  people = starwars
)
```

If you have a complicated directory system with nested templates and inheritance 
patterns, you may find it easier to initialize an `ExTera` by specifying the 
directory with a glob containing the `*` wildcard to indicate any number of 
template files.

```{r}
#| label: glob-example
template_dir <- file.path(tempdir(), "templates")

dir.create(template_dir)
dir.create(file.path(template_dir, "subdirectory"))

cat(
  '<p>Hello {{ x }}. This is {{ y }}.</p>',
  file = file.path(template_dir, "hello-world.html")
)

cat(
  '<h2>{{ title }}</h2>
<ol>
{%- for person in people %}
  {%- if person.films is containing("A New Hope") %}
  {%- if person.species and person.species is containing("Human") %}
  <li>{{ person.name }} ({{ person.homeworld }})</li>
  {%- endif %}
  {%- endif %}
{%- endfor %}
</ol>
',
  file = file.path(template_dir, "subdirectory", "star-wars.html")
)

glob <- file.path(template_dir, "*.html")

tera <- ExTera$new(glob)
tera

tera$render_to_string(
  "subdirectory/star-wars.html",
  title = "Humans of A New Hope",
  people = starwars
)
```